---
title: "Bio 110 Heart Rate lab"
output:
  pdf_document: default
  html_notebook: default
---

## Intro to R and RStudio Server

(flush this out.....)

## Coding with R

### Comments

One of the most important things you can do with any scripting language is to add comments to your code. Comments are basically instructions or descriptions you insert in your code that aren't executed as code. They're sometimes referred to as "love letters to your future self".

In R, a comment is preceded by a hash mark (#):

```{r}
# This is a comment. Use it to explain what's going on!

```

You don't have to comment every single command, but it's often a good idea to add enough comments so your future self (or collaborator) will understand your code.
 
### Assigning variables

One of the most common things to do in any scripting language is to create a variable and set it equal to something. In R, you use the "<-" (left-pointing arrow) to assign a value to a variable. The value can be an integer, text string, expression (e.g. "2+2"), or output of a function. Take a look at the example below:

```{r}
# assign a value to a:
a<-"my value for a"

# assign a value to myFavoriteNumber:
myFavoriteNumber<-27

# print a and myFavoriteNumber
a
myFavoriteNumber

```

Now it's your turn. In the code block below, set a variable "myVar" equal to five times 20, and then print it:

```{r}
#enter your code here


```

### Functions

Functions are basically shortcut commands that perform a certain task. Functions usually take input values (often called "arguments"), and return some output. Let's take a look below at the "sqrt" function:

```{r}
# sets "value" equal to the sqaure root of 20
value<-sqrt(20)

value

```

One of the great things about RStudio is that you can quickly access documentation on functions by using the "help" function. Running this function will display documentation in the lower right-hand window:

```{r}

help(sqrt)

```

Notice in the "Arguments" section, "x" is labled as "a numeric or complex vector or array". This is a pretty key piece of information, mainly because sometimes you may want to explicitly tell the function which argument you're sending to the function. This is especially true when the function accepts lots of arguments. If you don't explicitly define the argument, the function will assume you're sending arguments in the order desrcibed in the documentation. Take a look at this example:

```{r}
sqrt(9)

#explicitly defining arguments:
sqrt(x=9)

#this won't work:
# sqrt(y=9)

```

Why does the "x" work, but not the "y"? Because the documentation on sqrt defines the argument as "x".

Now it's your turn. Use help to look up the "plot" function. Then use the plot function, and explicitly label the x value 20, and the y value 30.

```{r}
#enter and run your "help" code here 


```


```{r}
#enter and run your "plot" code here


```


### Vectors

Vectors are a type of data structure in R, usually referring to an array of values. To create a vector, you use the "c" function:

```{r}
myFirstVector<-c(1,2,3,4,5)

#print myFirstVector
myFirstVector
```

Now it's your turn. Create a vector named "vector", and have it contain the values 30, 20, 40, 15, and 50.

```{r}
#enter your code here


```



### Plotting Heart Rate

Now let's apply some of these concepts and functions above to plot heart rate. We wish to create a simple scatterplot with continuous x and y variables, find the best-fit line, and determine the values for the slope and y-intercept. 

```{r}
#start by entering a vector of the x-values
x<-c(0, 20, 40, 60, 80, 100, 120)
#and one for the y-values (in order)
y1<-c(0, 24, 50, 70, 96, 121, 124)
#then plot y against x 
#the operator "~" means "value on the left depends upon value on the right"... or "y1 is a function of x": 
plot(y1~x)


```

We can add additional information to the plot, including specific x and y axis limits, axes labels, and a title.

```{r}
#if you want to adjust the axis limits, add labels, and a title

plot(y1~x, xlim=c(0,120), ylim=c(0,140), main="Greg's heart beats", xlab="time in seconds", ylab = "number of heartbeats")


```



```{r}
# in order to run the abline function below, you must first run the plot function:
plot(y1~x, xlim=c(0,120), ylim=c(0,140), main="Greg's heart beats", xlab="time in seconds", ylab = "number of heartbeats")

#to add the line, setting the y interect at zero, and get its slope, fit a linear model to the data and store the results in a variable (named "model" here).
model<-lm(formula=y1~0+x)
#add a best-fit line to the graph
abline (reg=model)
#type the name of the variable to see values of the slope and y-intercept
model
```


You can plot a second set of values by first creating another vector:

```{r}
#to add another set of points and another line...
y2<-c(0, 28, 58, 80, 100, 135, 140)


```

Now we can use the "matplot" function, which lets you plot multiple columns or sets of data:

```{r}

matplot(x=x, y=cbind(y1,y2), type="p", col=c("red", "green"), pch=c(16,16), main="Greg's heart rate increases with exercise", xlab="time in seconds", ylab="resting beats-red and exercising beats-green")

```




----- Last stuff below, explain better.


```{r}

plot(y1~x, xlim=c(0,120), ylim=c(0,140), main="Greg's heart beats", xlab="time in seconds", ylab = "number of heartbeats")


#repeat the linear model for y1
model<-lm(y1~0+x)
#plot its line in red
abline (model, col="red")
#and now the model for y2
model2<-lm(y2~0+x)
#and a green line for it
abline(model2, col="green")
#and to get the new slope and intercept
model2

#to add a legend, first need coordinates of the legend, c(“labels”), lty determines symbols with 1 = lines, lwd determines width of lines, colors 
legend(0,140, c("exercising heartrate", "resting heartrate"), lty=c(1,1), lwd=c(2.5, 2.5), col=c("green", "red") )


```


